---
title: "Development Guide: Adding Functions to WJPr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Development Guide: Adding Functions to WJPr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This technical guide walks through the complete process for adding new visualization functions to the WJPr package.

## Prerequisites

### Development Dependencies

```{r}
install.packages(c(
  "devtools",
  "roxygen2",
  "testthat",
  "usethis",
  "pkgdown"
))
```

### Clone the Repository

```{bash}
git clone https://github.com/worldjusticeproject/WJPr.git
cd WJPr
```

Then restore the project environment with `renv::restore()`.

## Step 1: Create the Function File

### Location and Naming

Create a new file in `R/` following the convention:

- **Charts**: `{type}Chart.R` (e.g., `waterfallChart.R`)
- **Utilities**: `{name}.R` (e.g., `validation.R`)

```{r}
# Create the file
usethis::use_r("waterfallChart")
```

## Step 2: Function Structure

### Complete Template

```{r}
#' Plot a Waterfall Chart following WJP style guidelines
#'
#' @description
#' `r lifecycle::badge("experimental")`
#'
#' `wjp_waterfall()` creates a waterfall chart showing the cumulative effect
#' of sequential positive and negative values.
#'
#' @param data Data frame containing the data to plot.
#' @param target String. Column name with numeric values to plot.
#' @param grouping String. Column name with category labels.
#' @param colors String. Column name for color grouping. Default is `NULL`.
#' @param cvec Named vector of colors. Default is `NULL` (the WJP palette,
#'   see [wjp_palette()], is applied).
#' @param labels String. Column name with value labels. Default is `NULL`.
#' @param direction String. `"horizontal"` or `"vertical"`. Default is `"vertical"`.
#' @param ptheme ggplot theme to apply. Default is [WJP_theme()].
#' @param show_legend Logical. If `TRUE`, displays a horizontal legend above
#'   the chart. Default is `FALSE`.
#'
#' @return A ggplot object representing the waterfall chart.
#' @export
#'
#' @examples
#' # Always load the WJP fonts
#' wjp_fonts()
#'
#' # Prepare data
#' data_waterfall <- data.frame(
#'   category = c("Start", "Gain A", "Loss B", "Gain C", "End"),
#'   value    = c(100, 30, -20, 15, 125),
#'   type     = c("total", "positive", "negative", "positive", "total")
#' )
#'
#' # Create chart
#' wjp_waterfall(
#'   data_waterfall,
#'   target   = "value",
#'   grouping = "category",
#'   colors   = "type",
#'   cvec     = c("total"    = "#482d8b",
#'                "positive" = "#0f9581",
#'                "negative" = "#f26b21")
#' )

wjp_waterfall <- function(
    data,
    target,
    grouping,
    colors      = NULL,
    cvec        = NULL,
    labels      = NULL,
    direction   = "vertical",
    ptheme      = WJP_theme(),
    show_legend = FALSE
) {

  # ===========================================================================
  # SECTION 1: INPUT VALIDATION (optional but recommended)
  # ===========================================================================

  if (!is.data.frame(data)) {
    stop("`data` must be a data frame.", call. = FALSE)
  }

  if (!target %in% names(data)) {
    stop("Column '", target, "' not found in `data`.", call. = FALSE)
  }

  if (!direction %in% c("vertical", "horizontal")) {
    stop('`direction` must be one of "vertical" or "horizontal".', call. = FALSE)
  }

  # ===========================================================================
  # SECTION 2: RENAME VARIABLES
  # ===========================================================================

  # Rename the required columns
  data <- data %>%
    dplyr::rename(
      target_var   = dplyr::all_of(target),
      grouping_var = dplyr::all_of(grouping)
    )

  # Handle the colors column
  if (is.null(colors) || identical(colors, grouping)) {
    # If not supplied (or the same column as grouping), reuse grouping
    data <- data %>%
      dplyr::mutate(colors_var = grouping_var)
  } else {
    data <- data %>%
      dplyr::rename(colors_var = dplyr::all_of(colors))
  }

  # Handle labels
  if (is.null(labels)) {
    data <- data %>%
      dplyr::mutate(labels_var = "")
  } else if (identical(labels, grouping)) {
    data <- data %>%
      dplyr::mutate(labels_var = grouping_var)
  } else if (identical(labels, colors)) {
    data <- data %>%
      dplyr::mutate(labels_var = colors_var)
  } else {
    data <- data %>%
      dplyr::rename(labels_var = dplyr::all_of(labels))
  }

  # ===========================================================================
  # SECTION 3: DATA TRANSFORMATIONS
  # ===========================================================================

  # Compute the waterfall positions
  data <- data %>%
    dplyr::mutate(
      end   = cumsum(target_var),
      start = dplyr::lag(end, default = 0),
      # Handle negative values
      ymin  = pmin(start, end),
      ymax  = pmax(start, end)
    )

  # ===========================================================================
  # SECTION 4: DEFAULT COLORS
  # ===========================================================================

  # Fall back to the WJP palette when no color vector is supplied
  if (is.null(cvec)) {
    cvec <- wjp_default_cvec(data$colors_var)
  }

  # ===========================================================================
  # SECTION 5: BASE CHART
  # ===========================================================================

  plt <- ggplot2::ggplot(
    data,
    ggplot2::aes(
      x    = grouping_var,
      fill = colors_var
    )
  ) +
    ggplot2::geom_rect(
      ggplot2::aes(
        xmin = as.numeric(factor(grouping_var)) - 0.4,
        xmax = as.numeric(factor(grouping_var)) + 0.4,
        ymin = ymin,
        ymax = ymax
      ),
      show.legend = show_legend
    ) +
    ggplot2::scale_fill_manual(
      values = cvec,
      breaks = wjp_legend_breaks(data$colors_var),
      name   = NULL
    )

  # Add value labels if supplied
  if (!all(data$labels_var == "")) {
    plt <- plt +
      ggplot2::geom_text(
        ggplot2::aes(
          x     = grouping_var,
          y     = ymax,
          label = labels_var
        ),
        vjust    = -0.5,
        family   = wjp_font_family(),
        fontface = "bold",
        size     = 3.514598,
        color    = "#4a4a49"
      )
  }

  # ===========================================================================
  # SECTION 6: DIRECTION
  # ===========================================================================

  if (direction == "horizontal") {
    plt <- plt +
      ggplot2::coord_flip()
  }

  # ===========================================================================
  # SECTION 7: THEME AND FINAL TWEAKS
  # ===========================================================================

  plt <- plt +
    ptheme +
    ggplot2::labs(x = "", y = "") +
    ggplot2::theme(
      panel.grid.major.x = ggplot2::element_blank()
    ) +
    wjp_legend_theme(show_legend)

  return(plt)
}
```

## Step 3: Important Patterns

### Handling Optional Parameters

The most important pattern is handling a parameter that is NULL or two parameters that point to the same column:

```{r}
# INCORRECT - Fails if colors == grouping
data <- data %>%
  rename(
    grouping_var = all_of(grouping),
    colors_var   = all_of(colors)  # ERROR: column already renamed
  )

# CORRECT - Check before renaming
if (is.null(colors) || identical(colors, grouping)) {
  data <- data %>%
    mutate(colors_var = grouping_var)
} else {
  data <- data %>%
    rename(colors_var = all_of(colors))
}
```

### Named Color Vector

```{r}
# Names must match the values of the colors variable EXACTLY
cvec <- c(
  "Trust"    = "#482d8b",
  "No Trust" = "#f26b21"
)

# Inside the function, fall back to the WJP palette when cvec is NULL
if (is.null(cvec)) {
  cvec <- wjp_default_cvec(data$colors_var)
}
plt <- plt +
  scale_fill_manual(values = cvec)
```

### Fonts

Never hard-code a font family. Read the active one with `wjp_font_family()` so `options(wjpr.family = "Inter Tight")` switches every chart:

```{r}
geom_text(
  family   = wjp_font_family(),
  fontface = "bold",
  size     = 3.514598  # 10 pt value labels
)
```

### Polar Coordinates (Circular Charts)

For radar, rose, or gauge charts:

```{r}
plt <- plt +
  coord_polar(
    theta = "y",      # Variable mapped to the angle
    start = pi        # Starting point (pi = bottom)
  ) +
  theme(
    aspect.ratio = 1  # Keep a 1:1 ratio for a circle
  )
```

### Invisible Segments

To control the visual range without affecting the data (as `wjp_gauge()` does to draw a semicircle):

```{r}
# Add a transparent padding segment
padding_data <- data.frame(
  colors_var = "___padding___",
  target_var = 50
)

data <- dplyr::bind_rows(data, padding_data)
cvec <- c(cvec, "___padding___" = "transparent")
```

Keep technical segments out of the legend by passing the visible categories as `breaks` (see `wjp_legend_breaks()`).

### Deprecating Parameters

When a parameter is renamed, keep the old name working and signal the change:

```{r}
wjp_waterfall <- function(data, target, grouping, order = NULL, order_var = deprecated()) {
  if (lifecycle::is_present(order_var)) {
    lifecycle::deprecate_soft("1.2.0", "wjp_waterfall(order_var)", "wjp_waterfall(order)")
    if (is.null(order)) order <- order_var
  }
  ...
}
```

## Step 4: Roxygen2 Documentation

### Required Tags

| Tag | Description | Example |
|-----|-------------|---------|
| `@description` | Detailed description | Include the lifecycle badge |
| `@param` | Every parameter | `@param data Data frame containing...` |
| `@return` | Returned value | `@return A ggplot object` |
| `@export` | Export the function | Always include |
| `@examples` | Reproducible example | Runnable code |

### Lifecycle Badge

Always include the function status:

```{r}
#' @description
#' `r lifecycle::badge("experimental")`
#'
#' Description of the function...
```

Options:
- `"experimental"` - New, may change
- `"stable"` - Stable API
- `"deprecated"` - Do not use, will be removed

## Step 5: Add Tests

Create or extend a file in `tests/testthat/`:

```{r}
test_that("wjp_waterfall returns a buildable ggplot", {
  data <- data.frame(
    category = c("Start", "Gain", "Loss"),
    value    = c(100, 30, -20)
  )

  plot <- wjp_waterfall(data, target = "value", grouping = "category")

  expect_s3_class(plot, "ggplot")
  expect_no_error(ggplot2::ggplot_build(plot))
})
```

## Step 6: Add a Visual Example

### Update generate-examples.R

Add a section to `data-raw/generate-examples.R`:

```{r}
# =============================================================================
# XX. WATERFALL CHART
# =============================================================================
message("XX. Generating waterfall chart example...")

data_waterfall <- data.frame(
  category = c("Start", "Revenue", "Costs", "Taxes", "End"),
  value    = c(100, 50, -30, -10, 110),
  type     = c("total", "gain", "loss", "loss", "total")
)

plot_waterfall <- wjp_waterfall(
  data_waterfall,
  target   = "value",
  grouping = "category",
  colors   = "type",
  cvec     = c("total" = "#482d8b",
               "gain"  = "#0f9581",
               "loss"  = "#f26b21")
)

save_example(plot_waterfall, "waterfall")
```

### Generate the Image

```{r}
source("data-raw/generate-examples.R")
```

## Step 7: Register the Function

- Add the function to `_pkgdown.yml` (the `functions` navbar menu and the `reference` index).
- Add a section to `vignettes/articles/gallery.Rmd`.
- Add an entry to `NEWS.md`.
- Add the function to the chart table in `CLAUDE.md`:

```markdown
| `wjp_waterfall()` | waterfallChart.R | Waterfall chart for cumulative changes |
```

## Step 8: Verification

### Pre-PR Checklist

```{r}
# 1. Regenerate documentation
devtools::document()

# 2. Run the tests
devtools::test()

# 3. Check the package
devtools::check()

# 4. Load and try it
devtools::load_all()
wjp_fonts()
# ... example code

# 5. Generate the image
source("data-raw/generate-examples.R")
```

### Common Errors

| Error | Cause | Solution |
|-------|-------|----------|
| `object 'x' not found` | Variable not renamed | Check the rename step |
| `Column doesn't exist` | Double rename | Use if/else to avoid it |
| `NULL` result | Missing `return()` | Add `return(plt)` |
| Font not found | Fonts not loaded | Call `wjp_fonts()` |
| `no visible binding for global variable` (check NOTE) | Column used in `aes()`/`mutate()` | Add it to `globalVariables()` in `R/WJPr-package.R` |

## Step 9: Pull Request

### Create a Branch

```{bash}
git checkout -b feature/waterfall-chart
```

### Commit

```{bash}
git add R/waterfallChart.R
git add man/wjp_waterfall.Rd
git add man/figures/example-waterfall.png
git add tests/testthat/test-charts.R
git add data-raw/generate-examples.R
git add _pkgdown.yml NEWS.md CLAUDE.md

git commit -m "Add wjp_waterfall() chart function

- New waterfall chart for cumulative changes
- Supports horizontal/vertical direction
- Full Roxygen2 documentation and tests
- Example image generated"
```

### Push and PR

```{bash}
git push origin feature/waterfall-chart
```

Open a PR on GitHub with:
- A description of the function
- A screenshot of the chart
- The completed checklist

## Additional Resources

- [R Packages (2e)](https://r-pkgs.org/) - Complete package development guide
- [ggplot2 extensions](https://exts.ggplot2.tidyverse.org/) - Inspiration
- [roxygen2 documentation](https://roxygen2.r-lib.org/) - Available tags
- [lifecycle](https://lifecycle.r-lib.org/) - Deprecating functions and arguments
