---
title: "Chart Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chart Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width  = 6,
  fig.height = 4,
  out.width  = "100%",
  dpi        = 150,
  fig.bg     = "white",
  dev.args   = list(bg = "white")
)
```

This gallery showcases the chart types available in WJPr using a consistent World Justice Project visual language: Lato typography, clear percentage labels, high-contrast text, restrained grid lines, and WJP brand colors. The examples use the package's sample data and are intended to demonstrate chart structure, not to reproduce findings from a specific report.

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(WJPr)

# Always load the WJP fonts before plotting
wjp_fonts()
```

## WJP Color Palette

The WJPr package uses colors from the WJP brand system:

| Color Name | Hex Code | Usage |
|------------|----------|-------|
| Violet (Primary) | `#482d8b` | Main series, section headers, emphasis |
| Teal-Blue | `#2894aa` | Secondary series and comparison groups |
| Orange | `#f26b21` | Contrast, warnings, negative or divergent values |
| Cool Gray | `#555659` | Body text, axes, notes, and supporting elements |

For categorical data visualizations, start with violet, teal-blue, and orange. When a chart needs more categories, use the established Rule of Law Index factor colors already included in this package. The full ordered palette is available programmatically via `wjp_palette()`:

```{r palette}
wjp_palette()
```

Every chart function falls back to this palette automatically when you do not supply a color vector (`cvec`), so quick exploratory charts stay on-brand by default.

## Typography

`wjp_fonts()` registers two font systems from Google Fonts: **Lato** (Full, Light, and Black weights) and **Inter Tight**. All charts use Lato by default. To switch every chart of the session — axis text, value labels, and legends — set the `wjpr.family` option once:

```{r typography, fig.height=3.5}
# Check the active family
wjp_font_family()

# Same chart, Inter Tight
options(wjpr.family = "Inter Tight")

wjp_bars(
  data.frame(
    country = c("Atlantis", "Narnia", "Neverland"),
    trust   = c(49, 46, 64)
  ) %>%
    mutate(label = paste0(trust, "%"), label_pos = trust + 6),
  target   = "trust",
  grouping = "country",
  labels   = "label",
  lab_pos  = "label_pos"
)

# Back to Lato (the default)
options(wjpr.family = NULL)
```

For a single chart, wrap the call with `withr::with_options(list(wjpr.family = "Inter Tight"), ...)`. To change only the theme elements (axis text and titles) of one chart, pass `ptheme = WJP_theme(family = "Inter Tight")` — note that value labels keep the session font in that case. See `wjp_font_family()` for details.

```{r data, message=FALSE, warning=FALSE}
# Load sample data
gpp_data <- WJPr::gpp
```

***
## Bar Charts

### Vertical Bars
`wjp_bars()` - Standard vertical bar chart for comparing values across categories.

```{r bars-vertical, warning=FALSE}
data_bars <- gpp_data %>%
  filter(year == 2022) %>%
  mutate(
    q1a  = as.double(unclass(q1a)),
    trust = case_when(q1a <= 2 ~ 1, q1a <= 4 ~ 0)
  ) %>%
  group_by(country) %>%
  summarise(trust = mean(trust, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(label = paste0(round(trust, 0), "%"), label_pos = trust + 5)

wjp_bars(
  data_bars,
  target   = "trust",
  grouping = "country",
  colors   = "country",
  labels   = "label",
  lab_pos  = "label_pos",
  cvec     = c("Atlantis" = "#482d8b", "Narnia" = "#2894aa", "Neverland" = "#f26b21")
)
```

**Expected input structure** — one row per category, with the value column (`target`) and the category column (`grouping`); `labels`/`lab_pos` are optional helpers for the data labels.

For stacked or multi-series bars, set `show_legend = TRUE` when `colors` identifies the segments or series. A legend is intentionally omitted for simple category bars like this example because it would repeat the axis labels.

```{r bars-structure, echo=FALSE}
knitr::kable(head(data_bars, 3), caption = "Input data for wjp_bars(): target = trust, grouping = country")
```

### Horizontal Bars
Set `direction = "horizontal"` for horizontal orientation.

```{r bars-horizontal, warning=FALSE}
wjp_bars(
  data_bars,
  target    = "trust",
  grouping  = "country",
  colors    = "country",
  labels    = "label",
  lab_pos   = "label_pos",
  cvec      = c("Atlantis" = "#482d8b", "Narnia" = "#2894aa", "Neverland" = "#f26b21"),
  direction = "horizontal"
)
```

Uses the same `data_bars` structure shown above — only `direction = "horizontal"` changes.

***

## Diverging Bars
`wjp_divbars()` - Show positive and negative values extending from a center point.

```{r divbars, warning=FALSE}
data_divbars <- gpp_data %>%
  filter(year == 2022) %>%
  mutate(
    q1a     = as.double(unclass(q1a)),
    response = case_when(q1a <= 2 ~ "Trust", q1a <= 4 ~ "No Trust")
  ) %>%
  filter(!is.na(response)) %>%
  group_by(country, response) %>%
  count() %>%
  group_by(country) %>%
  mutate(
    percent = (n / sum(n)) * 100,
    label   = paste0(round(percent, 0), "%")
  )

wjp_divbars(
  data_divbars,
  target    = "percent",
  grouping  = "country",
  diverging = "response",
  negative  = "No Trust",
  labels    = "label",
  cvec      = c("Trust" = "#482d8b", "No Trust" = "#f26b21"),
  show_legend = TRUE
)
```

**Expected input structure** — one row per `grouping` × `diverging` combination. The group named in `negative` is flipped into the negative quadrant automatically, so all values can be supplied as positive percentages. With `show_legend = TRUE`, the legend identifies the response groups using the names in `cvec`.

```{r divbars-structure, echo=FALSE}
knitr::kable(head(data_divbars, 4), caption = "Input data for wjp_divbars(): target = percent, grouping = country, diverging = response")
```

***

## Dots Chart
`wjp_dots()` - Compare multiple variables across groups with dot markers.

```{r dots, warning=FALSE}
data_dots <- gpp_data %>%
  select(country, q1a, q1b, q1c, q1d) %>%
  mutate(
    across(starts_with("q1"), \(x) as.double(unclass(x))),
    across(starts_with("q1"), ~ case_when(.x <= 2 ~ 1, .x <= 4 ~ 0))
  ) %>%
  group_by(country) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE) * 100), .groups = "drop") %>%
  pivot_longer(-country, names_to = "variable", values_to = "trust") %>%
  mutate(
    institution = case_when(
      variable == "q1a" ~ "Police",
      variable == "q1b" ~ "Courts",
      variable == "q1c" ~ "Parliament",
      variable == "q1d" ~ "Government"
    )
  )

wjp_dots(
  data_dots,
  target   = "trust",
  grouping = "institution",
  colors   = "country",
  cvec     = c("Atlantis" = "#482d8b", "Narnia" = "#2894aa", "Neverland" = "#f26b21"),
  show_legend = TRUE
)
```

**Expected input structure** — long format: one row per `grouping` (variable) × `colors` (group) combination, each with its `target` value. Set `show_legend = TRUE` to identify the groups mapped through `colors`.

```{r dots-structure, echo=FALSE}
knitr::kable(head(data_dots, 4), caption = "Input data for wjp_dots(): target = trust, grouping = institution, colors = country")
```

***

## Line Chart
`wjp_lines()` - Display trends over time with connected points.

```{r lines, warning=FALSE, message=FALSE}
library(ggrepel)

data_lines <- gpp_data %>%
  filter(country == "Atlantis") %>%
  select(year, q1a, q1b, q1c) %>%
  mutate(
    across(starts_with("q1"), \(x) as.double(unclass(x))),
    across(starts_with("q1"), ~ case_when(.x <= 2 ~ 1, .x <= 4 ~ 0)),
    year = as.character(year)
  ) %>%
  group_by(year) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE) * 100), .groups = "drop") %>%
  pivot_longer(-year, names_to = "variable", values_to = "trust") %>%
  mutate(
    institution = case_when(
      variable == "q1a" ~ "Police",
      variable == "q1b" ~ "Courts",
      variable == "q1c" ~ "Parliament"
    ),
    label = paste0(round(trust, 0), "%")
  )

wjp_lines(
  data_lines,
  target   = "trust",
  grouping = "year",
  colors   = "institution",
  labels   = "label",
  repel    = TRUE,
  cvec     = c("Police" = "#482d8b", "Courts" = "#2894aa", "Parliament" = "#f26b21"),
  show_legend = TRUE
)
```

**Expected input structure** — long format: one row per `grouping` (time point) × series. The `colors` variable defines the lines, so single- and multi-series data use the same call. Use `show_legend = TRUE` to label those series with a shared legend.

```{r lines-structure, echo=FALSE}
knitr::kable(head(data_lines, 4), caption = "Input data for wjp_lines(): target = trust, grouping = year, colors = institution")
```

***

## Slope Chart
`wjp_slope()` - Compare values between exactly two time points.

```{r slope, warning=FALSE}
data_slope <- gpp_data %>%
  filter(year %in% c(2017, 2019)) %>%
  mutate(
    q1a   = as.double(unclass(q1a)),
    gend  = as.double(unclass(gend)),
    trust  = case_when(q1a <= 2 ~ 1, q1a <= 4 ~ 0),
    gender = case_when(gend == 1 ~ "Male", gend == 2 ~ "Female")
  ) %>%
  group_by(year, gender) %>%
  summarise(trust = mean(trust, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(label = paste0(round(trust, 0), "%"))

wjp_slope(
  data_slope,
  target   = "trust",
  grouping = "year",
  colors   = "gender",
  labels   = "label",
  cvec     = c("Male" = "#482d8b", "Female" = "#f26b21"),
  repel    = TRUE,
  show_legend = TRUE
)
```

**Expected input structure** — exactly two `grouping` values (the two time points) per series. The `colors` variable defines the lines, and `show_legend = TRUE` identifies them above the chart.

```{r slope-structure, echo=FALSE}
knitr::kable(head(data_slope, 4), caption = "Input data for wjp_slope(): target = trust, grouping = year, colors = gender")
```

***

## Dumbbell Chart
`wjp_dumbbells()` - Show change between two points with connected markers.

```{r dumbbells, warning=FALSE}
data_dumbbells <- data_lines %>%
  filter(year %in% c("2017", "2022"))

wjp_dumbbells(
  data_dumbbells,
  target   = "trust",
  grouping = "institution",
  colors   = "year",
  cgroups  = c("2017", "2022"),
  labels   = "label",
  cvec     = c("2017" = "#2894aa", "2022" = "#482d8b"),
  show_legend = TRUE,
  label_offset = 4
)
```

**Expected input structure** — one row per `grouping` × `colors` (the two endpoints). `cgroups` lists the two endpoint values and supplies the legend labels. The optional `labels` column adds a color-matched value label next to each point; `label_offset` controls its distance from the endpoint.

```{r dumbbells-structure, echo=FALSE}
knitr::kable(head(data_dumbbells, 4), caption = "Input data for wjp_dumbbells(): target = trust, grouping = institution, colors = year")
```

***

## Lollipop Chart
`wjp_lollipops()` - Minimalist bar alternative with stems and dots.

```{r lollipops, warning=FALSE}
wjp_lollipops(
  data_bars,
  target   = "trust",
  grouping = "country"
)
```

Uses the same `data_bars` structure (one row per category with a `target` value) shown under **Bar Charts**. Value labels are generated automatically; pass a `labels` column to override them.

***

## Edgebars Chart
`wjp_edgebars()` - Horizontal bars with labels at the edge, ideal for narrow spaces.

```{r edgebars, warning=FALSE}
wjp_edgebars(
  data_bars,
  target   = "trust",
  grouping = "country",
  cvec     = "#2894aa"
)
```

Uses the same `data_bars` structure shown under **Bar Charts**. The text drawn at the edge of each bar defaults to the `grouping` values; pass a `labels` column to customize it.

***

## Radar Chart
`wjp_radar()` - Compare multiple dimensions on a circular grid.

```{r radar, fig.height=5, warning=FALSE}
data_radar <- gpp_data %>%
  select(gend, starts_with("q49")) %>%
  mutate(
    gend = as.double(unclass(gend)),
    across(starts_with("q49"), \(x) as.double(unclass(x))),
    gender = case_when(gend == 1 ~ "Male", gend == 2 ~ "Female"),
    across(starts_with("q49"), ~ case_when(.x <= 2 ~ 1, .x <= 99 ~ 0))
  ) %>%
  group_by(gender) %>%
  summarise(across(starts_with("q49"), ~ mean(.x, na.rm = TRUE) * 100), .groups = "drop") %>%
  pivot_longer(-gender, names_to = "category", values_to = "score") %>%
  mutate(
    label = case_when(
      category == "q49a"    ~ "Reliable",
      category == "q49b_G1" ~ "Accessible",
      category == "q49b_G2" ~ "Channels",
      category == "q49c_G1" ~ "Consistent",
      category == "q49c_G2" ~ "Stable",
      category == "q49d_G1" ~ "Efficient",
      category == "q49d_G2" ~ "Fast",
      category == "q49e_G1" ~ "Effective",
      category == "q49e_G2" ~ "Trustworthy"
    )
  )

wjp_radar(
  data_radar,
  axis_var = "category",
  target   = "score",
  labels   = "label",
  colors   = "gender",
  cvec     = c("Male" = "#482d8b", "Female" = "#f26b21"),
  show_legend = TRUE
)
```

**Expected input structure** — long format: one row per `axis_var` (dimension) × `colors` (group). `labels` gives the human-readable axis name, while `show_legend = TRUE` identifies the compared groups.

```{r radar-structure, echo=FALSE}
knitr::kable(head(data_radar, 4), caption = "Input data for wjp_radar(): target = score, axis_var = category, colors = gender")
```

***

## Rose Chart
`wjp_rose()` - Circular bar chart for single-unit multi-dimensional data.

```{r rose, fig.height=5, warning=FALSE}
data_rose <- data_radar %>%
  filter(gender == "Male")

wjp_rose(
  data_rose,
  target   = "score",
  grouping = "category",
  labels   = "label",
  cvec     = wjp_palette(9)
)
```

**Expected input structure** — a single unit: one row per `grouping` (dimension) with its `target` value and a display `label`.

```{r rose-structure, echo=FALSE}
knitr::kable(head(data_rose, 4), caption = "Input data for wjp_rose(): target = score, grouping = category")
```

***

## Gauge Chart
`wjp_gauge()` - Semicircular chart for showing composition or progress.

```{r gauge, warning=FALSE}
data_gauge <- data.frame(
  category = c("Factor 1", "Factor 2", "Factor 3", "Factor 4"),
  value    = c(25, 35, 25, 15),
  label    = c("25%", "35%", "25%", "15%")
)

wjp_gauge(
  data_gauge,
  target       = "value",
  colors       = "category",
  cvec         = c("Factor 1" = "#482d8b", "Factor 2" = "#2894aa",
                   "Factor 3" = "#f26b21", "Factor 4" = "#555659"),
  factor_order = c("Factor 1", "Factor 2", "Factor 3", "Factor 4"),
  labels       = "label",
  show_legend  = TRUE
)
```

**Expected input structure** — one row per slice: a `colors` category, its `target` value, and a display `label`. `factor_order` controls the drawing order, and `show_legend = TRUE` identifies each colored slice.

```{r gauge-structure, echo=FALSE}
knitr::kable(head(data_gauge, 4), caption = "Input data for wjp_gauge(): target = value, colors = category")
```

***

## Grouped Bars
`wjp_groupbars()` - Faceted bars that compare a value across demographic groups (one facet per `grouping`). Values can be supplied as proportions (`0-1`) or percentages (`0-100`). Confidence intervals can be calculated from `sd` and `sample_size`, or supplied directly with `ci_lower` and `ci_upper`. A national/general value can be added either as its own bar (`national_style = "bar"`) or as a vertical reference line (`national_style = "line"`).

```{r groupbars, fig.height=5, fig.retina=1, fig.bg="white", warning=FALSE}
# Percentage-scale input with precomputed confidence intervals.
data_groupbars <- data.frame(
  group    = c("Gender", "Gender", "Age", "Age", "Age"),
  category = c("Men", "Women", "18-24", "25-54", "55+"),
  value    = c(74.4, 70.3, 72.1, 73.1, 73.0),
  lower    = c(72.4, 68.4, 70.1, 71.0, 70.8),
  upper    = c(76.4, 72.1, 74.5, 75.2, 75.2)
)

wjp_groupbars(
  data_groupbars,
  target            = "value",
  grouping          = "group",
  levels            = "category",
  colors            = c("#482d8b", "#e5e8e8"),
  group_order       = c("Gender", "Age"),
  level_order       = list(
    Gender = c("Men", "Women"),
    Age    = c("18-24", "25-54", "55+")
  ),
  draw_ci           = TRUE,
  ci_lower          = "lower",
  ci_upper          = "upper",
  show_national     = TRUE,
  national_value    = 72.3,
  national_style    = "bar",
  national_label    = "National Average",
  national_ci_lower = 70.0,
  national_ci_upper = 74.6,
  show_axis         = TRUE
)
```

**Expected input structure** — one row per `levels` category, tagged with the `grouping` (facet) it belongs to and its `target` value. `target`, `national_value`, `ci_lower`, and `ci_upper` may be proportions or percentages. To draw a confidence interval on each bar, set `draw_ci = TRUE` and either supply per-category `sd` plus `sample_size`, or pass precomputed `ci_lower` and `ci_upper` columns. When `national_style = "bar"`, the national value is inserted as a regular bar row and can receive its own `national_ci_lower` and `national_ci_upper`. The second color represents the complement to 100%, so a neutral gray keeps attention on the measured percentage and its interval.

```{r groupbars-structure, echo=FALSE}
knitr::kable(head(data_groupbars, 5), caption = "Input data for wjp_groupbars(): target = value, grouping = group, levels = category, lower/upper for confidence intervals")
```

***

## Quick Reference

| Chart | Function | Best For |
|-------|----------|----------|
| Vertical Bars | `wjp_bars()` | Comparing values across categories |
| Horizontal Bars | `wjp_bars(direction = "horizontal")` | Long category names |
| Diverging Bars | `wjp_divbars()` | Positive/negative comparisons |
| Dots | `wjp_dots()` | Multiple groups, multiple variables |
| Lines | `wjp_lines()` | Trends over time |
| Slope | `wjp_slope()` | Change between two time points |
| Dumbbells | `wjp_dumbbells()` | Before/after comparisons |
| Lollipops | `wjp_lollipops()` | Minimalist bar alternative |
| Edgebars | `wjp_edgebars()` | Narrow spaces, long labels |
| Radar | `wjp_radar()` | Multi-dimensional group comparison |
| Rose | `wjp_rose()` | Multi-dimensional single unit |
| Gauge | `wjp_gauge()` | Composition, progress |
| Grouped Bars | `wjp_groupbars()` | Compare a value across demographic groups |
